####

#
#Project Report for Practical Machine Learning Course                                                                  


This is file created to produce the report during home work assignment of  practical Machine learning Coursera project.
The scripts has been produced  and tested in R Studio.


###Backround introduction


Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 




### Goal of Project





The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases. 




####Data 


The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 




###Reproduceablity
 
 In order to reproduce the same  code add user has to load the libraries into the code file and download the above Data into your working directory.

### Load the required Libraries 


```{r}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)


set.seed(1234)
```


## Loading the data and preliminary clean up the data


```{r}

# Some missing values are coded as string "#DIV/0!" or "" or "NA" - these will be changed to NA.
# We notice that both data sets contain columns with all missing values - these will be deleted.  

trainingSet <- read.csv("pml-training.csv",na.strings=c("NA","#DIV/0!",""))

testSet <- read.csv("pml-testing.csv" ,na.strings=c("NA","#DIV/0!",""))


#check the dimentions for both the datas

dim(trainingSet)
dim(testSet)


##take out columns with missing values.
trainingset<-trainingSet[,colSums(is.na(trainingSet)) == 0]
testingset <-testSet[,colSums(is.na(testSet)) == 0]

#The features user_name raw_timestamp_part_1 raw_timestamp_part_2 cvtd_timestamp
#new_window num_window are not related to calculations and are removed form the
#downloaded data.

trainingset   <-trainingset[,-c(1:7)]
testingset <-testingset[,-c(1:7)]


dim(trainingset)
dim(testingset)

```


##Partitioning the training data set to allow cross-validation

The training data set contains 53 variables and 19622 obs.
The testing data set contains 53 variables and 20 obs.
In order to perform cross-validation, the training data set is partionned into 2 sets: subTraining (75%) and subTest (25%).
This will be performed using random subsampling without replacement.


```{r}
subsamples <- createDataPartition(y=trainingset$classe, p=0.75, list=FALSE)
subTraining <- trainingset[subsamples, ] 
subTesting <- trainingset[-subsamples, ]
dim(subTraining)
dim(subTesting)
#head(subTraining)
#head(subTesting)


```



## Bar plot for classe variables


The variable "classe" contains 5 levels: A, B, C, D and E. A plot of the outcome variable will allow us to see the frequency of each levels in the subTraining data set and compare one another.

```{r}
plot(subTraining$classe, col="yellow", main="Bar Plot of levels of the variable classe within the subTraining data set", xlab="classe levels", ylab="Frequency")

```

From the graph it is seen that each level of frequency is within the same order. Level A is the most frequent with more than 4000 occurence and level D has the least frequency 2500 occurence.



###Prediction model with Decision 


```{r}
model1 <- rpart(classe ~ ., data=subTraining, method="class")
#model1
# Predicting:
prediction1 <- predict(model1, subTesting, type = "class")
#prediction1
# Plot of the Decision Tree
rpart.plot(model1, main="Classification Tree", extra=102, under=TRUE, faclen=0)

#Test result on subtesting data set
confusionMatrix(prediction1, subTesting$classe)


```

###Second prediction model: Using Random Forest

```{r}
model2 <- randomForest(classe ~ . , data=subTraining, method ="class")
prediction2 <- predict(model2, subTesting, type=  "class")
confusionMatrix(prediction2,subTesting$classe)
```

### Decision

Accuracy for Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is choosen. The accuracy of the model is 0.995. The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.



### Submission


```{r}
predictfinal <- predict(model2, testingset, type="class")
predictfinal

# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)

```
